Joint Likelihood Deconvolution of Astronomical Images in the Presence of Poisson Noise

### 2023-09-26 src/tex/ms.tex
[showyourwork] overleaf sync

% Abstract with filler text
    \begin{abstract}
        We present a new method for Joint Likelihood Deconvolution (Jolideco) of a set of astronomical observations of the same sky region in the presence of Poisson noise. The method reconstructs a single flux image from a set of observations by optimizing the \aposteriori joint Poisson likelihood of all observations under a patch based image prior. The patch prior is parameterised by a Gaussian Mixture model (GMM) which we trained on astronomical images with high signal to noise ratio, including data from the  James Webb Telescope as well as the GLEAM radio survey. During the reconstruction process the patch prior adapts to the patch structures in the data by finding the most likely GMM component for each patch in the image. By applying the method to simulated data we show that both the combination of multiple observations as well as the patch based prior lead to a much improved reconstruction quality in many different source scenarios as well as signal to noise regimes. We show also that the method yields superior reconstruction quality to alternative standard methods such as the Richardson-Lucy method. We also the results of the method applied to example data from the \chandra observatory as well as the \fermi. In both cases \jolideco improves the angular resolution as well as signal to noise ratio 
        by a factor of 42 \todo{it would be good to summarize the results with a few key numbers...}.
    \end{abstract}

@@

    To maximise the scientific use of available data it is desirable to correct for the instrumental imprint on the data as well as reduce its noise level. The process of improving the resolution of an image in post-processing is often called deblurring or deconvolution, as it aims to invert the effect of the data being degraded by a forward convolution through the point spread function (PSF) of the instrument. In general this class of problems is categorised as \textit{inverse problems}. If there is prior information available on the shape of the PSF, we typically refer to the process more precisely as unblind deconvolution. In case of astronomical observations the PSF can for example be estimated empirically from observation of known point sources, such as distant activate galactic nuclei or a model of the PSF can obtained from simulation of the telescope or instrument.

    \subsection{Related Work}
    In literature there have been many efforts to solve the unblind deconvolution problem for images affected by Poisson noise. Most methods rely fundamentally on modeling the image generating process as a linear process. Given a known point spread function a prediction of the discrete measured image can be obtained via \textit{forward folding} and be compared against an observation assuming Poisson statistics per pixel. However the methods differ in their additional prior or regularisation assumptions as well as their approach to optimization or sampling of the underlying likelihood function.

    The standard baseline method for image deconvolution was proposed independently by \cite{Richardson1972} and \cite{Lucy1974} (we will further refer to this method as \textit{RL}). The method involves estimating the original image by iteratively convolving the observed image with the PSF and dividing by the result of convolving the estimated image with the transposed PSF. The iterations continue until the estimated image converges to a satisfactory solution.

    However in practice the RL method was found to have multiple, partly related limitations: first there is no objective criterion for when to stop the optimization process. Usually the iteration is stopped when the reconstructed image seems plausible to the eye. Second, when the number of iterations is too high, the method shows a tendency to decompose regions of constant brightness into a set of bright emission peaks (\enquote{speckles}). This problem was e.g. described and studied by \cite{Reeves1995} and \cite{Fish1995}. Lastly the RL method does not provided an associated estimate of the uncertainties of the reconstruction, which is desirable especially in the Poisson domain to measure the significance of specific features found in the reconstructed image.

@@

    To address the issue with RL decomposing extended emission structures into groups of point source, multiple additions to the RL method were proposed. Most importantly this lead to the introduction of regularised RL methods, such as use of total variation \citep{Dey2006} or sparseness \todo{add reference}. The challenge is due to fundamentally contradicting goals: deconvolution favoring point sources will decompose extend structures, while introducing local correlations will lead to less sharp reconstrcution of point sources.

    Multi-scale LIRA prior has been used in image deconvolution to incorporate error estimates and Bayesian sampling. This prior has been used in various studies, including those by \cite{Esch2004} and \cite{Connors2011}. However, the prior is ad-hoc and lacks physical information or assumptions, limiting its applicability.

@@

    \cite{Pumpe2018} and \cite{Selig2015} approached the problem of deconvolution of Poisson noise data implicitly using the framework of information field theory. Deconvolution, Denoising and Decomposition (D3PO) and (D4PO).

@@

    Interesting approaches to the problem of deconvolution can of course be found in the field of computer vision. For example \cite{Zoran2011} build on very similar assumption of local correlation for natural images. Instead of assuming a fixed local correlation structure, such as smoothness defined by a Gaussian with a correlation size, they proposed to learn the local correlation structure on millions of patches extracted from a set of reference images. This approach can be seen as a form of transfer learning. For this they proposed to split training images from representative data sets into small patches of size 8x8~pixels and fit a 64 dimensional Gaussian Mixture Model (GMM) to the distribution of extracted patches. Each pixel of the patch 
    is thereby treated as an independent dimension in the model. In total
    they chose to use $k=200$ Gaussian components. To use the GMM in reconstruction they introduce the expected patch likelihood (EPLL). They showed that the patch based GMM prior led to much improved image reconstructions for multiple inverse problems such as denoising, deblurring and inpainting. \cite{Bouman2016} later adapted the patch prior successfully for the reconstruction of radio data from the Event Horizon Telescope (EHT).

@@

    \subsection{Some Remarks on Deep Learning Methods}
    In the past decade Machine learning (ML) especially deep convolutional neural networks (CNN) became the \textit{de facto} standard for deblurring and super-resolution of natural images \todo{reference review?}. However their success relies on the availability of massive sets of training data with typically low noise and corresponding ground truth to perform supervised learning. For a variety of reasons, which we will outline in the following, this approach is not ideal for astronomical x-ray and \gammaray data.

@@

    In addition to the challenge of obtaining training data, it is known that the PSF of instruments such as \chandra is highly variable. It varies with observation conditions and depends e.g. on energy as well as offset angle from the pointing direction. To cover the whole parameter space of the PSF one would need to train a network for each of the PSF models.

@@

    \begin{itemize}
        \item Different observations of one instrument or telescope at different times and observation conditions. E.g., multiple observations of \chandra of the same astrophysical object with different offset angles and exposure times
        \item Observations from different telescopes, which operate in the same wavelength range. E.g., a \chandra and \xmm observation of the same region in the sky
        \item A single observation of one telescope with different data quality categories and different associated instrument response functions, such as event classes for \fermi
    \end{itemize}

@@

    Using Maximum Likelihood Estimation (MLE) we could try to get an estimate of $\hat{\mathbf{x}}$ directly, using some optimization procedure. However in general this represents an \textit{ill posed} inverse problem. Estimates for $\mathbf{x}$ which have similar likelihood values $\mathcal{L}(\hat{\mathbf{x}}_1)$ look very different from each other and might also look different compared to a given ground truth, if available. Both the convolution operation as well as the exitence of noise lead to a loss of information in the data which cannot easily be recovered. The likelihood surface shows many local, close-by minima. Depending on the initial estimate for $\mathbf{x}$ and the minimization algorithm the MLE based estimate is likely to only find such a local minimum. This is fundamentally the reason for RL decomposing estimates into point sources, described in the previous section.

    \subsection{A Posteriori Joint Likelihood}
    When reconstructing images in the context of inverse problems we are operating in a high dimensional parameter space, because each pixel in the image represents an independent parameter in the optimization process. By using prior information we can introduce correlations between the parameters, to reduce the \textit{effective} dimensionality of the problem and guide the optimization process towards unique and more stable solutions.

    Using Bayes rule we can estimate the \aposteriori likelihood under a given prior:
    %
    \begin{equation}
        \label{eq:bayes}
        P(\mathbf{x}|\textbf{D}) = P(\mathbf{x} ) \frac{P(\textbf{D} |\mathbf{x})}{P(\textbf{D})}
    \end{equation}

@@

    Where $\mathcal{C}\left( \mathbf{D_j} | \mathbf{x} \right)$ represents the summed log-likelihood for an individual observation $j$. Additionally the function includes a log-prior term $\mathcal{P}(\mathbf{x})$ and a factor $\beta$ to adjust the weight of the prior with respect to the joint log likelihood term. The parameter $\beta$ can be seen as a hyper-parameter, chosen by the user to quantify their belief in the correctness of the chosen prior distribution. \todo{this is somewhat dissatisfying...but in practice very common...see e.g. any kind of regularisation...if there is enough data cross validation could be used to tune this hyper-parameter}.

    \subsection{Patch Based Image Prior}
    In general it is difficult to capture global image statistics in a prior likelihood taking into account all pixel to pixel correlations. The size of the corresponding correlation matrix would grow with the number of pixels squared and thus quickly becomes computationally untraceable. However astronomical images typically only show local correlations between pixels. This is because the spatial scale of astrophysical processes in an image is always limited by the finite interaction time and the fact that the interaction cannot propagate faster than the speed of light (structure in the CMB might be an exception here). This means on smaller scales astronomical images often contain basic structures such as edges, corners, filaments or periodic patterns formed by local astrophysical processes. For this reason we adopt the idea of a patch based image prior approach as introduced by \cite{Zoran2011}. Given a pre-trained Gaussian Mixture Model (GMM) for the distribution of pixel values in patches, we can evaluate a patch based prior for the reconstructed image $\mathbf{x}$ and approximate the \aposteriori likelihood.

@@

    Where $\mathbf{P}_n$ is a matrix that extracts the $n$-th patch from the image $\mathbf{x}$ to be reconstructed. This way the index $k$ that identifies the most likely GMM component per patch, becomes a discrete hyper-parameter in the model that is optimized along with the image $\mathbf{x}$. The bottom row of Figure~\ref{fig:patches} shows an example of a patch and the same patch represented in the \textit{Eigenbasis} of the covariance matrices of its three most likely GMM components. Visually the representation of the of the patch in the covariance \textit{Eigenbasis} works as a denoising of the patch.

@@

    With dividing by the mean of the patch $\Bar{\mathbf{x}}_n$ the structure in the patch becomes independent of the total brightness the patch contains. This results in patches being grouped only by their pixel to pixel correlation structure and not by total contrast. This corresponds the assumption that the structure in the patch is statistically independent of the total flux contained in the patch, which we consider a valid assumption for astronomical images. This also generalizes to the prior likelihood of the whole image described by Equation~\ref{eq:prior-total}. The value of the prior likelihood becomes independent of the total flux contained in the image. The prior just regularizes the structure in the image, while the overall intensities are typically well constrained by the Poisson likelihood term.

@@

    % The choice of the image normalisation allows to adjust the contrast of the image
    % and e.g. enhance low intensity structures, which also enhances the regularising
    % effect of the patch prior.

@@

    % There are arguments for $\arcsinh(\mathbf{x})$, see e.g. \cite{Lupton2004}, which we use.

    \subsubsection{Cycle Spinning}
    To avoid artifact due to the choice of the grid of overlapping patches we propose three variations:

    \begin{itemize}
        \item Cycle spinning by randomly shifting the image by a given number of pixels in x and y direction, e.g. used in \cite{Esch2004}
        \item Sub pixel cycle spinning, by randomly distributing the brightness of a given pixel to a 4 pixel grid \todo{This is something I came up with to allow for point sources to have sub-pixel positions, but I never showed it actually works...}
        \item A randomly chosen patch grid, such that each pixel is at least covered once. The random grid of patches was proposed by \cite{Parameswaran2018}. This has the advantage of potentially speeding up the computation as well. \todo{It is available in the implementation, but a bit ad hoc, maybe just not mention...}
    \end{itemize}

@@

    Cross validation is feasible once sufficient data is available, which is e.g. the 
    case for deep \chandra~observations.

@@

    \subsection{Systematic Errors of Predicted Counts}
    Each measured counts image $\mathbf{D_j}$ is associated with systematic errors. This includes e.g. the pointing accuracy of the telescopes: when repeatedly pointed to the same object or when using different types of telescopes for observation, the absolute position of an astronomical object might vary in the image between different observations. The process of correcting for these inaccuracies is called \textit{astrometry} and a standard procedure in for example \chandra data analysis. A second example of systematic errors is the variation of background emission between multiple observations.

@@

    \section{Implementation}
    \subsection{Jolideco Framework}
    The goal of the reconstruction process is to optimize the a-posteriori likelihood defined by Equation~\ref{eq:total}. Given that each pixel in the reconstructed image represents an independent parameter this represents a high dimensional optimization problem. Differentiable programming frameworks such as \texttt{PyTorch}~\citep{Pytorch2019} allow for solving these kind of high dimensional modeling problems, by using back-propagation and adapted optimization methods for high dimensional models such as stochastic gradient decent.

    We implemented the \jolideco method as an independent Python package, based on \texttt{PyTorch} as optimization back-end. We tried to define a modular, object oriented code structure, to allow parts of the algorithm to be flexible and interchangeable, such as choice of the patch normalisation, the choice of GMM models, optimization methods and serialisation formats for the reconstruction results as well as corresponding diagnostic information such as the trace of the posterior and prior likelihood values. We hope that this simplifies the future improvement and extension of the package. The package is available at \url{https://github.com/jolideco/jolideco}.

    We also use \texttt{Numpy}~\citep{Numpy2020} for handling data arrays and \texttt{Matplotlib}~\citep{Hunter2007} for plotting. To learn the Gaussian Mixture Models from patches we use the standard implementation provided in the \texttt{Scikit-Learn}~\citep{Skimage2014} package. To handle the serialisation of images to the \textit{FITS} data format and to handle world coordinate systems (WCS) we use \texttt{Astropy}~\citep{Astropy2018}.

@@

    \subsubsection{JWST Cassiopeia A}
    As a specific prior for the analysis of SNRs we learned a GMM prior from data of the James Webb Space Telescope (JWST).
    The JWST is a large, infrared space telescope launched in late 2021. It delivers image data at very high spatial resolution as well as excellent signal to noise ratio. Both of these properties make JWST data an excellent choice for extracting patches for training a GMM. To limit the data reduction effort we relied on a promotional image of the SNR \textit{Cassiopeia A}\footnote{We downloaded the full resolution (4k)  PNG image from \url{https://webbtelescope.org/contents/media/images/2023/121/01GWQBBY77MHGFV3M3N63KDCEJ}}.
    We first converted the image to gray scale using a perceived brightness model \todo{reference skimage...}. Then we split the image into $\approx 2$ million patches and learned a GMM with 128 mixture components. Again we relied on the patch normalisation described in Section~\ref{ssec:patch-norm}. Figure~\ref{fig:gmm-eigen-images} shows the weighted sum of the \textit{Eigenimages} of the covariance matrices of the first 32 components of the GMM. The patches show a large variety in morphological shapes, including edges, gradients, bright compact spots and more combinations thereof.

@@

    % We first split the compiled image into 150 individual examples. For each image we ran a blob detection algorithm and line fit to find the preferred direction of the jet. Using this information we aligned all images to the same horizontal reference direction of the jet split all images into patches of size 8x8 pixels. In total we extracted \todo{insert number} patches and used those to train a GMM with 32 components. The prior is useful in a scenario were the direction of the jet known from e.g. previous radio observations.

@@

    \subsection{Optimization Strategy}
    Maximising the \aposteriori likelihood defined Eq.\ref{eq:total} with respect to the model parameters $\mathbf{x}$ is a high dimensional optimization problem. To improve the convergence rate of the optimization process we employ the following optimization strategy: the optimization strategy begins with the optimizer viewing $\log{\mathbf{x}}$, and the first step involves running a rough estimate using a uniform prior. This step is optional, but if chosen, the next step involves optimizing the parameters associated with the systematic error described in Eq.~\ref{eq:model-npred-calibration}. Following this, a full optimization is performed using the patch prior, which can be computationally expensive. The optimization algorithm used is ADAM, a stochastic gradient descent method, without decay. Finally, there is an optional step of learning rate scheduling, although so far, a fixed step size value of $\mathrm{lr}=10^{-3}$ has been used.

@@

    On scenario A1 all methods show acceptable to excellent results for the isolated point sources in the upper right of the image. This demonstrates all methods take into account the PSF and background models correctly. The early stopped uniform prior still shows an extension for those cases, because the optimization process has not finished. Optimizing for 1000 iterations clearly improves the result and shows almost perfect point sources corresponding to the size of approximately one pixel. The \jolideco method with the Zoran-Weiss patch prior results in less sharp images, because the prior has been learned on natural images of landscapes and every day objects and scenes, which do not include point sources. The \jolideco method with the prior learned from GLEAM data, which also includes point sources, shows much improved results. However not as sharp as the results obtained with the uniform prior. In the lower left of the image the point sources become fainter and move closer, challenging all methods in succeeding to separate those as independent sources. The uniform prior optimized \endquote{to end} shows excellent results, still separating the sources.

    For scenario B3 all method recover the point sources and the central extend source. However large difference can be observed in the reconstruction of the Gaussian shaped extended sources. As expected the uniform prior decompose the emission into many point sources, neglecting the spatial correlation between them. \textbf{Pylira} shows slightly improved results with the gaps between the bright emission peeks being more filled, however still lacks a convincing result into reconstructing the source as a single object. The \jolideco method with the patch priors shows the best results both visually as well as measured by the SSI. The GLEAM v0.1 patch prior reconstructs the sources almost perfectly, visually almost indistinguishable from the ground truth. The Zoran \& Weiss prior also shows good reconstruction of the extended sources, but less smooth and leaving a visible extension in the surrounding point sources.

    Scenario C3 challenges the methods to reconstruct a weak extended source under four brighter point sources. While again all methods reconstruct the point sources well, the visual differences are large for the weak underlying extended emission. The uniform prior with 1000 iterations decomposes the source into a grid of almost equidistant point sources resembling only the outer boundary of the original source. \textit{Pylira} shows a similar tendency but already improves the reconstruction of the extended source by leaving more emission in between the bright emission spots. The best results are achieved by the patch priors, which reconstruct a plausible extended source with approximately uniform brightness.

    On the jet like features the patch prior does not show a preferred direction or bias in reconstruction and preserves the linear correlation between the pixels.

>>>>

% Abstract with filler text
    \begin{abstract}
        We present a new method for Joint Likelihood Deconvolution (Jolideco) of a set of astronomical observations of the same sky region in the presence of Poisson noise. The method reconstructs a single flux image from a set of observations by optimizing the \aposteriori joint Poisson likelihood of all observations under a patch based image prior. The patch prior is parameterised by a Gaussian Mixture model (GMM) which we trained on astronomical images with high signal to noise ratio, including data from the  James Webb Telescope as well as the GLEAM radio survey. During the reconstruction process the patch prior adapts to the patch structures in the data by finding the most likely GMM component for each patch in the image. By applying the method to simulated data we show that both the combination of multiple observations as well as the patch based prior lead to a much improved reconstruction quality in many different source scenarios as well as signal to noise regimes. We show also that the method yields superior reconstruction quality to alternative standard methods such as the Richardson-Lucy method. We also show the results of the method applied to example data from the \chandra observatory as well as the \fermi. 
    \end{abstract}

@@

    To maximise the scientific use of available data it is desirable to correct for the instrumental imprint on the data as well as reduce its noise level. The process of improving the resolution of an image in post-processing is often called \enquote{deblurring} or \enquote{deconvolution}, as it aims to invert the effect of the data being degraded by a forward convolution through the point spread function (PSF) of the instrument. In general this class of problems is categorised as \enquote{inverse problems}. If there is prior information available on the shape of the PSF, we typically refer to the process more precisely as \enquote{unblind deconvolution}. In case of astronomical observations the PSF can for example be estimated empirically from observation of known point sources, such as distant activate galactic nuclei or a model of the PSF can obtained from simulation of the telescope or instrument.

    \subsection{Related Work}
    In literature there have been many efforts to solve the unblind deconvolution problem for images affected by Poisson noise. Most methods rely fundamentally on modeling the image generating process as a linear process. Given a known PSF a prediction of the discrete measured image can be obtained via \enquote{forward folding} and be compared against an observation assuming Poisson statistics per pixel. However the methods differ in their additional prior or regularisation assumptions as well as their approach to optimization or sampling of the underlying likelihood function. In the following section we give a short overview over those methods.

    The standard baseline method for image deconvolution was proposed independently by \cite{Richardson1972} and \cite{Lucy1974} (we will further refer to this method as \enquote{\textit{RL}}). Their method involves the

    involves treating the forward convolution by the PSF in a Bayesian and solving the e

    The method involves estimating the original image by iteratively convolving the observed image with the PSF and dividing by the result of convolving the estimated image with the transposed PSF. The iterations continue until the estimated image converges to a satisfactory solution.

    However in practice the RL method was found to have multiple, partly related limitations: first there is no objective criterion for when to stop the optimization process. Usually the iteration is stopped when the reconstructed image seems plausible to the eye. Second, when the number of iterations is too high, the method shows a tendency to decompose regions of constant brightness into a set of bright emission peaks (\enquote{speckles}). This problem was for example described and studied by \cite{Reeves1995} and \cite{Fish1995}. Lastly the RL method only provide a \enquote{point estimate} and no associated estimate of the uncertainties of the reconstruction, which is desirable especially in the Poisson domain for example to measure the significance of specific features found in the reconstructed image.

@@

@@

    To address the issue with RL decomposing extended emission structures into groups of point source, multiple additions to the RL method were proposed. Most importantly this lead to the introduction of regularised RL methods, such as use of total variation \citep{Dey2006} or sparse representations based on \enquote{wavelets} or \enquote{curvelets} \citep{Starck2003}. The challenge is rooted in two fundamentally contradicting goals: a deconvolution process that favors point sources will decompose extend structures present in the image, while introducing local correlations between neighbouring pixels will lead to less sharp reconstruction of point sources.

    One possible approach to overcome this challenge has been proposed by \cite{Esch2004} and \cite{Connors2011} as the \enquote{Low Counts Image Reconstrction and Analysis} (LIRA) method. Their key idea was to introduce a prior assumption that favors smoothness on multiple spatial scales of the image. Instead of a trying to find a single point estimate for the reconstructed image  they also changed to sampling from the posterior distribution, which allowed them to quantify the associated uncertainties. However the prior distribution is based on the generic assumption of smoothness and does not encode any \enquote{real world} or physical information on the structures present in the image.

     \cite{Selig2015} and \cite{Pumpe2018} approached the problem of deconvolution of Poisson noise data implicitly using the framework of information field theory. Deconvolution, Denoising and Decomposition (D3PO) and (D4PO).

@@

    Other interesting approaches to the problem of deconvolution can of course be found in the field of computer vision. For example \cite{Zoran2011} build on very similar assumption of local correlation for natural images. However instead of assuming a fixed local correlation structure, such as smoothness defined by a Gaussian with a correlation size, they proposed to learn the local correlation structure on millions of patches extracted from a set of reference images. This approach can be seen as a form of transfer learning. For this they proposed to split training images from representative data sets into small patches of size 8x8~pixels and fit a 64 dimensional Gaussian Mixture Model (GMM) to the distribution of extracted patches. Each pixel of the patch  is thereby treated as an independent dimension in the model. In total they chose to use $k=200$ Gaussian components. To use the GMM in reconstruction they introduce the expected patch likelihood (EPLL). They showed that the patch based GMM prior led to much improved image reconstructions for multiple inverse problems such as denoising, deblurring and inpainting. \cite{Bouman2016} later adapted the patch prior successfully for the reconstruction of radio data from the Event Horizon Telescope (EHT).

@@

    \subsection{Some Remarks on Deep Learning Methods}
    In the past decade Machine learning (ML) especially deep convolutional neural networks (CNN) became the standard for deblurring and super-resolution of natural images \todo{reference review?}. However their success relies on the availability of massive sets of training data with typically low noise and corresponding ground truth to perform supervised learning. For a variety of reasons, which we will outline in the following, this approach is not ideal for astronomical x-ray and \gammaray data.

@@

    In addition to the challenge of obtaining training data, it is known that the PSF of instruments such as \chandra is highly variable. It varies with observation conditions and depends for example on energy as well as offset angle from the pointing direction. To cover the whole parameter space of the PSF one would need to train a network for each of the PSF models.

@@

    \begin{itemize}
        \item Different observations of one instrument or telescope at different times and observation conditions. For example, multiple observations of \chandra of the same astrophysical object with different offset angles and exposure times
        \item Observations from different telescopes, which operate in the same wavelength range. For example, a \chandra and \xmm observation of the same region in the sky
        \item A single observation of one telescope with different data quality categories and different associated instrument response functions, such as event classes for \fermi
    \end{itemize}

@@

    Using Maximum Likelihood Estimation (MLE) we could try to get an estimate of $\hat{\mathbf{x}}$ directly, using some optimization procedure. However in general this represents an \textit{ill posed} inverse problem. Estimates for $\mathbf{x}$ which have similar likelihood values $\mathcal{L}(\hat{\mathbf{x}}_1)$ look very different from each other and might also look different compared to a given ground truth, if available. Both the convolution operation as well as the existence of noise lead to a loss of information in the data which cannot easily be recovered. The likelihood surface shows many local, close-by minima. Depending on the initial estimate for $\mathbf{x}$ and the minimization algorithm the MLE based estimate is likely to only find such a local minimum. This is fundamentally the reason for RL decomposing estimates into point sources, described in the previous section.

    \subsection{A Posteriori Joint Likelihood}
    When reconstructing images in the context of inverse problems we are operating in a high dimensional parameter space, because each pixel in the image represents an independent parameter in the optimization process. By using prior information we can introduce correlations between the parameters, to reduce the \enquote{effective dimensionality} of the problem and guide the optimization process towards unique and more stable solutions.

    Using Bayes' rule we can estimate the \aposteriori likelihood under a given prior:
    %
    \begin{equation}
        \label{eq:bayes}
        P(\mathbf{x}|\textbf{D}) = P(\mathbf{x} ) \frac{P(\textbf{D} |\mathbf{x})}{P(\textbf{D})}
    \end{equation}

@@

    Where $\mathcal{C}\left( \mathbf{D_j} | \mathbf{x} \right)$ represents the summed log-likelihood for an individual observation $j$. Additionally the function includes a log-prior term $\mathcal{P}(\mathbf{x})$ and a factor $\beta$ to adjust the weight of the prior with respect to the joint log likelihood term. The parameter $\beta$ can be seen as a hyper-parameter, chosen by the user to quantify their belief in the correctness of the chosen prior distribution. The choice of the parameter is optional and for all experiments we show later it is simply fixed to $\beta=1$.

    \newpage
    \subsection{Patch Based Image Prior}
    In general it is difficult to capture global image statistics in a prior likelihood taking into account all pixel to pixel correlations. The size of the corresponding correlation matrix would grow with the number of pixels squared and thus quickly becomes computationally untraceable. However astronomical images typically only show local correlations between pixels. This is because the spatial scale of astrophysical processes in an image is always limited by the finite interaction time and the fact that the interaction cannot propagate faster than the speed of light (structure in the CMB might be an exception here). This means on smaller scales astronomical images often contain basic structures such as edges, corners, filaments or periodic patterns formed by local astrophysical processes. For this reason we adopt the idea of a patch based image prior approach as introduced by \cite{Zoran2011}. Given a pre-trained Gaussian Mixture Model (GMM) for the distribution of pixel values in patches, we can evaluate a patch based prior for the reconstructed image $\mathbf{x}$ and approximate the \aposteriori likelihood.

@@

    Where $\mathbf{P}_n$ is a matrix that extracts the $n$-th patch from the image $\mathbf{x}$ to be reconstructed. This way the index $k$ that identifies the most likely GMM component per patch, becomes a discrete hyper-parameter in the model that is optimized along with the image $\mathbf{x}$. The bottom row of Figure~\ref{fig:patches} shows an example of a patch and the same patch represented in the \enquote{Eigenbasis} of the covariance matrices of its three most likely GMM components. Visually the representation of the of the patch in the Eigenbasis of the covariance results in a denoised version of the patch.

@@

    With dividing by the mean of the patch $\Bar{\mathbf{x}}_n$ the structure in the patch becomes independent of the total brightness the patch contains. This results in patches being grouped only by their pixel to pixel correlation structure and not by total contrast. This corresponds to the assumption that the structure in the patch is statistically independent of the total flux contained in the patch, which we consider a valid assumption for astronomical images. This also generalizes to the prior likelihood of the whole image described by Equation~\ref{eq:prior-total}. The value of the prior likelihood becomes independent of the total flux contained in the image. The prior just regularizes the structure in the image, while the overall intensities are typically well constrained by the Poisson likelihood term.

@@

    % The choice of the image normalisation allows to adjust the contrast of the image
    % and for example enhance low intensity structures, which also enhances the regularising
    % effect of the patch prior.

@@

    % There are arguments for $\arcsinh(\mathbf{x})$, see for example \cite{Lupton2004}, which we use.

    \subsubsection{Cycle Spinning}
    The evaluation of the patch prior divides the reconstructed image into a fixed grid of overlapping patches. To avoid artifacts in the reconstruction due to the fixed position of the grid, we use the method of \enquote{cycle spinning}. Before evaluating the patch prior we randomly shift the image $\mathbf{x}$ by a given number of pixels in x and y direction. Were the shift is smaller than half the size of a patch. The same method has, for example, been used in \cite{Esch2004}. We also experimented with cycle spinning by a sub-pixel amount, mostly for the reconstruction of point sources. However we later introduced the possibility to reconstruct an upsampled version of $\mathbf{x}$, which made the sub-pixel cycle-spinning mostly redundant.

    A variation of cycle spinning specifically for the patch prior was proposed by \cite{Parameswaran2018}. Instead of a regular grid for the patches they proposed to work with a randomised grid of patches, where the center of each patch was randomly shifted, while ensuring that patches still overlap by at least one pixel. In doing so they found they could work with a smaller total number of patches, while retaining the same reconstruction quality. This improved the computational performance of the method. We also implemented the evaluation of the patch prior on a randomized patch grid for comparison and found similar results as \cite{Parameswaran2018}: the reconstruction was slightly faster with the same reconstruction quality. However for all results shown later in the manuscript we chose the simple cycle spinning with upsampling.

@@

    Cross validation is feasible once sufficient data is available, which is for example the 
    case for deep \chandra~observations.

@@

    \subsection{Systematic Errors of Predicted Counts}
    Each measured counts image $\mathbf{D_j}$ is associated with systematic errors. This includes for example the pointing accuracy of the telescopes: when repeatedly pointed to the same object or when using different types of telescopes for observation, the absolute position of an astronomical object might vary in the image between different observations. The process of correcting for these inaccuracies is called \textit{astrometry} and a standard procedure in for example \chandra data analysis. A second example of systematic errors is the variation of background emission between multiple observations.

@@

    \section{Implementation}
    \subsection{Jolideco Framework}
    The goal of the reconstruction process is to optimize the a-posteriori likelihood defined by Equation~\ref{eq:total}. Given that each pixel in the reconstructed image represents an independent parameter this represents a high dimensional optimization problem. Differentiable programming frameworks such as \texttt{PyTorch}~\citep{Pytorch2019} allow for solving these kind of high dimensional modeling problems, by using back-propagation and adapted optimization methods for high dimensional models such as stochastic gradient decent. The latter is also applicable in our case, as we work with multiple observations at the same time, where a subset of those represents a \enquote{batch} in terms of ML.

    We implemented the \jolideco method as an independent Python package, based on \texttt{PyTorch} as optimization back-end. We tried to define a modular, object oriented code structure, to allow parts of the algorithm to be flexible and interchangeable, such as choice of the patch normalisation, the choice of GMM models, optimization methods and serialisation formats for the reconstruction results as well as corresponding diagnostic information, such as the trace of the posterior and prior likelihood values. We hope that this simplifies the future improvement and extension of the package as well as the adoption by the astromical community. The package is available at \url{https://github.com/jolideco/jolideco}.

    We also use \texttt{Numpy}~\citep{Numpy2020} for handling data arrays and \texttt{Matplotlib}~\citep{Hunter2007} for plotting. To learn the Gaussian Mixture Models from patches we use the standard implementation provided in the \texttt{Scikit-Learn}~\citep{scikit-learn} package. To handle the serialisation of images to the \textit{FITS} data format and to handle world coordinate systems (WCS) we use \texttt{Astropy}~\citep{Astropy2018}.

@@

    \subsubsection{JWST Cassiopeia A}
    As a specific prior for the analysis of SNRs we learned a GMM prior from data of the James Webb Space Telescope (JWST).
    The JWST is a large, infrared space telescope launched in late 2021. It delivers image data at very high spatial resolution as well as excellent signal to noise ratio. Both of these properties make JWST data an excellent choice for extracting patches for training a GMM. To limit the data reduction effort we relied on a promotional image of the SNR \textit{Cassiopeia A}\footnote{We downloaded the full resolution (4k)  PNG image from \url{https://webbtelescope.org/contents/media/images/2023/121/01GWQBBY77MHGFV3M3N63KDCEJ}}.
    We first converted the image to gray scale using a perceived brightness model \citep{scikit-image}. Then we split the image into $\approx 2$ million patches and learned a GMM with 128 mixture components. Again we relied on the patch normalisation described in Section~\ref{ssec:patch-norm}. Figure~\ref{fig:gmm-eigen-images} shows the weighted sum of the \textit{Eigenimages} of the covariance matrices of the first 32 components of the GMM. The patches show a large variety in morphological shapes, including edges, gradients, bright compact spots and more combinations thereof.

@@

    % We first split the compiled image into 150 individual examples. For each image we ran a blob detection algorithm and line fit to find the preferred direction of the jet. Using this information we aligned all images to the same horizontal reference direction of the jet split all images into patches of size 8x8 pixels. In total we extracted \todo{insert number} patches and used those to train a GMM with 32 components. The prior is useful in a scenario were the direction of the jet known from for example previous radio observations.

@@

    \subsection{Optimization Strategy}
    Maximising the \aposteriori likelihood defined Eq.\ref{eq:total} with respect to the model parameters $\mathbf{x}$ is a high dimensional optimization problem.

    To improve the convergence rate of the optimization process we employ the following optimization strategy: the optimization strategy begins with the optimizer viewing $\log{\mathbf{x}}$, and the first step involves running a rough estimate using a uniform prior. This step is optional, but if chosen, the next step involves optimizing the parameters associated with the systematic error described in Eq.~\ref{eq:model-npred-calibration}. Following this, a full optimization is performed using the patch prior, which can be computationally expensive. The optimization algorithm used is ADAM, a stochastic gradient descent method, without decay. Finally, there is an optional step of learning rate scheduling, although so far, a fixed step size value of $\mathrm{lr}=10^{-3}$ has been used.

@@

    On scenario A1 all methods show acceptable to excellent results for the isolated point sources in the upper right of the image. This demonstrates all methods take into account the PSF and background models correctly. The early stopped uniform prior still shows an extension for those cases, because the optimization process has not finished. Optimizing for 1000 iterations clearly improves the result and shows almost perfect point sources corresponding to the size of approximately one pixel. The \jolideco method with the Zoran-Weiss patch prior results in less sharp images, because the prior has been learned on natural images of landscapes and every day objects and scenes, which do not include point sources. The \jolideco method with the prior learned from GLEAM data, which also includes point sources, shows much improved results. However not as sharp as the results obtained with the uniform prior. In the lower left of the image the point sources become fainter and move closer, challenging all methods in succeeding to separate those as independent sources. The uniform prior optimized \endquote{to end} shows good results, still separating the sources. All other methods cannot fully resolve weak and close by point sources. \jolideco with th e

    For scenario B3 all methods recover the point sources well. However large difference can be observed in the reconstruction of the Gaussian shaped extended sources. As expected, the uniform prior assumption lets the emission decomose into many point sources, neglecting the spatial correlation between them. \textbf{Pylira} shows slightly improved results with the gaps between the bright emission peeks being more filled, however still lacks a convincing result of reconstructing the source as a single object. The \jolideco method with the patch priors shows the best results both visually as well as measured by the SSI. The GLEAM v0.1 patch prior reconstructs the sources almost perfectly, visually almost indistinguishable from the ground truth. The Zoran \& Weiss prior also shows good reconstruction of the extended sources, but less smooth and leaving a more visible extension in the surrounding point sources.

    Scenario C3 challenges the methods to reconstruct a weak extended source under four brighter point sources. While again all methods reconstruct the point sources well, the visual differences are large for the weak underlying extended emission. The uniform prior with 1000 iterations decomposes the source into a grid of almost equidistant point sources resembling only the outer boundary of the original source. \textit{Pylira} shows a similar tendency but already improves the reconstruction of the extended source by leaving more emission in between the bright emission spots. The best results are achieved by the patch priors, which reconstruct a plausible extended source with approximately uniform brightness. The jet like features in the same scenario are decomposed into a \enquote{thread} of point sources by the uniform prior and LIRA method. The patch prior achieves a much more plausible reconstruction with almost uniform brightness for the linear structures. We also note that the patch prior does not show a preferred direction or bias in reconstruction, that is possibly encoded in the training data.
